Storm surges in the Strait of Georgia simulated with a regional model

### 2015-05-13 AO_guides/manuscriptAO.tex
Worked on revisions

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed. It is shown that the surge entering the domain from the Pacific Ocean contributes most significantly to the variability. Comparisons between simulations with high resolution and low resolution atmospheric forcing further emphasize this point. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas during a major wind storm on Dec 15, 2006. Generally, surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca.
\end{abstract}

@@

The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait (Figure \ref{fig:domain}). At its southern end, the Strait of Georgia is connected to the Strait of Juan de Fuca through several passages, the largest being Haro Strait. The San Juan and Gulf Islands are scattered throughout these passages where complex bathymetry and steep topographic slopes induce vigorous tidal mixing.  Mixed diurnal and semi-diurnal tides are found throughout most of the Salish Sea but a semi-diurnal amphidrome is present nearby Victoria, where the tides are mainly diurnal \citep{thomson1981oceanography}. This system is home to several large coastal communities, such as Vancouver and Victoria, serves as a waterway for commercial traffic and recreational use, and supports a large and active fishing industry.

@@

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk of flooding and property damage caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  There is also a small contribution to increased sea level due to positive sea surface height anomalies propagating northward from the equator along the eastern Pacific coast during El Nino years \citep{strub2002altimeter}. Communities in the Strait of Georgia are particularly susceptible to storm surges in the winter months when there is an increased propensity for storms and a higher mean sea level caused by southeasterly winds and geostrophic adjustment \citep{danard2003storm}. In the summer months, the wind direction changes to northwesterly and mean sea level is lower. In British Columbia waters, the largest tidal ranges occur in June and December at the summer and winter solstices \citep{thomson1981oceanography}. The highest water levels are observed in the winter around the perigean spring tide or ``King" tide, when the sun and moon are aligned with the Earth and the moon is at its perigee. Storm surges in the Strait of Georgia are typically observed between November and February \citep{forseth2006adaptation}. Flooding caused by a storm surge is most likely to occur during a high spring tide \citep{abeys2011extreme}.

Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{mercer2002barotropic, weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructured grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion surface waves \citep{xu2010storm} have also been identified as important.  Most of these models neglect the effects of stratification by treating the ocean as a constant temperature and salinity. An accurate forecast of the meteorological conditions at the ocean surface is of course essential in the generation of useful storm surge forecasts.

%storm surge modelling
Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} who used a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters \citep{murty1995storm, danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting, bernier2007tide, bernier2010tide}, relatively little has been published on storm surge modelling on the west coast of Canada in recent years. However, a storm surge forecasting program is currently operating over a domain that covers the entire British Columbia coastline \citep{StormBC}. Further, a statistical overview of past surge events in British Columbia waters suggests an increase in extreme water level events during warm El Nino periods \citep{abeys2011extreme}.

%overview oF SoG/Salish Sea models
The early modelling work of \citet{crean1988mathematical}, \citet{crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. %A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}.
These models pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, typically from 0.002 to 0.004 in the baroclinic model, especially in the vicinity of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too low and the M$_2$ tidal phases were too large in the Strait of Georgia \citep{crean1988mathematical}. Local adjustments to bottom friction were applied in both their barotropic and baroclinic models.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction although their domain did not include Puget Sound and the northern Strait of Georgia. \citet{foreman2004m} assimilated tide gauge harmonics to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia and noted a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. 
%A finite volume model on an unstructured grid by \citet{yang2010multi} involved a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. 
\citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. They noted that their semi-diurnal amplitudes were generally too low. An updated version of this model produced high surface temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. Smaller domains have also been considered; notably, the passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling this region.

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its skill in simulating storm surge hindcasts. To complement previous barotropic storm surge modelling in this region, this model employs a higher resolution and includes the effect of stratification and rivers. However, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. That section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's skill in producing storm surge hindcasts is assessed in section \ref{sec:storm}. Several simulations comparing the effects of different forcing conditions are also discussed. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modular tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study.

@@

The curvilinear orthogonal numerical grid has been divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels were stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10 m the grid was stretched to a maximum grid spacing of 27 m in the lowest layer. At the bottom boundary, partial $z$-levels were used in order to limit large changes in bathymetry across grid cells \citep{madec2012nemo}. {\color{red} Model constraints require that all passages be at least two grid points wide.  This restriction required choices in either widening or closing narrow straits.  In particular, Active Pass was closed and Porlier Pass was widened.  Shallow regions were represented by constant depth of 4 m.  All of these compromises are expected to have significant effects locally but minimal effects on the large scale.}

@@

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results of a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid to give a two-dimensional forcing that varies with time.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. {\color{red}Tidal predictions were generated with a tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge. In order to prevent over-predicted constituents due to high surge years, the time series is first filtered by applying a Doodson filter (NOAA citation) and removing periods with large non-tidal amplitudes. Long period constituents such as SA, SSA, etc are not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected. } The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

{\color{red} Although Tofino is not the closest tide guage to the western boundary, we have found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform. }

\begin{table}[h]
\centering 
\tbl{Average surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the surge events discussed on this article. A negative delay indicates that the peak surge occurred later than at Tofino.} 
{\begin{tabular}{|l |c c} 
\hline
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
Tofino  & 0.707 & -- \\
Toke Point & 1.17 & -2.83 \\
Neah Bay & 0.658 & -1.83 \\
Bamfield & 0.658 & -0.5 \\
Winter Harbour & 0.572 & 0.333 \\
Port Hardy & 0.491 & -2.67
\hline
\end{tabular}}
\end{table}
The tidal forcing and remote forcing were applied using the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}, a radiation scheme that allows disturbances to propagate out of the domain without reflections off of the open boundary. The barotropic velocity component of the remote forcing was set to zero. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4 (the version used here).  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. At the north baroclinic velocities were set to zero.

@@

The ocean surface is forced with momentum and heat fluxes from a 33-km global atmospheric reforecasting model suitable for use in ocean modelling \citep{smith2014new}. Forecasts from the period of 2002-2012 are available. One additional simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the surface atmospheric pressure is also included in the ocean momentum equations \citep{madec2012nemo}. In the open ocean, the surface atmospheric pressure primarily induces an inverse barometer (IB) sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the acceleration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure (set to 101,000 Pa). Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. In coastal and shelf seas, the atmospheric pressure forcing can also induce significant non-IB sea surface height, as demonstrated in previous studies of the Hudson Bay and Labrador shelf system \citep{wright1987influence,young1995synoptic}.

River input provides a significant volume of fresh water to the Salish Sea and can influence stratification, circulation, and primary productivity. However, most rivers in the domain are not gauged so parametrizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages. Freshwater runoff from each watershed was divided among the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source from 0 to 3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface temperature. A total of 150 rivers were parametrized by this method and their locations are indicated by the green dots in Figure \ref{fig:domain}.

Initial conditions for temperature and salinity came from a CTD cast in the middle Strait of Georgia taken in Sept 2002 \citep{pawlowiczetal2007}. Conditions were initially uniform horizontally and velocity was initialized at zero. The model was spun up for 15.5 months from the initial conditions above, starting Sept 16, 2002, using atmospheric forcing from 2002-2003, climatological temperature and salinity and sea surface height at the boundaries, with tides and climatological river output.  All storm surge runs were started at least three days prior to the event of interest with zero initial velocities and sea surface height and a stratification profile from model spin up. The modelled sea surface height adjusted to forcing in less than one day {\color{red} as determined by comparing a simulation with restart conditions to one with velocities and sea surface height initialized to zero}.

>>>>

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through {\color{red} the} comparisons of tidal elevation harmonics and storm surge with observations. Important forcing factors contributing to storm surges have been assessed. It is shown that the surge entering the domain from the Pacific Ocean contributes most significantly to the {\color{red} surge amplitude within the Strait of Georgia. Comparisons between simulations with high resolution and low resolution atmospheric forcing further emphasize tthat remote forcing is the dominant factor in surge amplitudes of this region.} In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas during a major wind storm on Dec 15, 2006. Generally, surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca.
\end{abstract}

@@

The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia {\color{red}\citep{thomson1981oceanography}}. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait (Figure \ref{fig:domain}). At its southern end, the Strait of Georgia is connected to the Strait of Juan de Fuca through several passages, the largest being Haro Strait. The San Juan and Gulf Islands are scattered throughout these passages where complex bathymetry and steep topographic slopes induce vigorous tidal mixing {\color{red}\citep{thomson1981oceanography}}.  Mixed diurnal and semi-diurnal tides are found throughout most of the Salish Sea but a semi-diurnal amphidrome is present nearby Victoria, where the tides are mainly diurnal \citep{thomson1981oceanography}. This system is {\color{red} a} home to several large coastal communities, such as Vancouver and Victoria, serves as a waterway for commercial traffic and recreational use, and supports a large and active fishing industry.

@@

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk of flooding and property damage caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  There is also a small contribution to increased sea level due to positive sea surface height anomalies propagating northward from the equator along the eastern Pacific coast during El Nino years \citep{strub2002altimeter}. Communities in the Strait of Georgia are particularly susceptible to storm surges in the winter months when there is an increased propensity for storms and a higher sea level caused by southeasterly winds and geostrophic adjustment \citep{danard2003storm}. In the summer months, the wind direction changes to northwesterly and mean sea level is lower. In British Columbia waters, the largest tidal ranges occur in June and December at the summer and winter solstices \citep{thomson1981oceanography}. The highest water levels are observed in the winter around the perigean spring tide or ``King" tide, when the sun and moon are aligned with the Earth and the moon is at its perigee. Storm surges in the Strait of Georgia are typically observed between November and February \citep{forseth2006adaptation}. Flooding caused by a storm surge is most likely to occur during a high spring tide \citep{abeys2011extreme}.

Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{mercer2002barotropic, weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructured grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion {\color{red} of} surface waves \citep{xu2010storm} have also been identified as important.  Most of these models neglect the effects of stratification by treating the ocean as a constant temperature and salinity. An accurate forecast of the meteorological conditions at the ocean surface is of course essential in the generation of useful storm surge forecasts.

%storm surge modelling
Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} who used a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters \citep{murty1995storm, danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting, bernier2007tide, bernier2010tide, ma2015oceanic}, relatively little has been published on storm surge modelling on the west coast of Canada in recent years. However, a storm surge forecasting program is currently operating over a domain that covers the entire British Columbia coastline \citep{StormBC}. Further, a statistical overview of past surge events in British Columbia waters suggests an increase in extreme water level events during warm El Nino periods \citep{abeys2011extreme}.

%overview oF SoG/Salish Sea models
The early modelling work of \citet{crean1988mathematical}, \citet{crean1988numerical} and \citet{stronach1993update} {\color{red} developed} a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. %A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}.
These models pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, typically from 0.002 to 0.004 in the baroclinic model, especially in the vicinity of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too low and the M$_2$ tidal phases were too large in the Strait of Georgia \citep{crean1988mathematical}. Local adjustments to bottom friction were applied in both their barotropic and baroclinic models.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction although their domain did not include Puget Sound and the northern Strait of Georgia. \citet{foreman2004m} assimilated tide gauge harmonics to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia and noted a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. 
%A finite volume model on an unstructured grid by \citet{yang2010multi} involved a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. 
\citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. They noted that their semi-diurnal amplitudes were generally too low. {\color{red} An updated version of this model with added biogeochemical components and an extended domain produced higher than observed surface temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}.} Smaller domains have also been considered; notably, the passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling this region.

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its skill in simulating storm surge hindcasts. {\color{red} The main purposes are to identify the forcing factors that contribute most significantly to storm surges in the Strait of Georgia, to examine the spatial distribution of the storm surge, and to evaluate the performance of a new model of this region.}  To complement previous barotropic storm surge modelling in this region, this model employs a higher resolution and includes the effect of stratification and rivers. However, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. That section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's skill in producing storm surge hindcasts is assessed in section \ref{sec:storm}. Several simulations comparing the effects of different forcing conditions are also discussed. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modular tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. {\color{red} Although these features are not being used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia.} This work focuses on the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study. {\color{red} NEMO has been successfully used in other coastal modelling studies in the Northeast Atlantic \citep{maraldi2013nemo} and Mediterranean (CITATION)!}

@@

The curvilinear orthogonal numerical grid has been divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels were stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10 m the grid was stretched to a maximum grid spacing of 27 m in the lowest layer. At the bottom boundary, partial $z$-levels were used in order to limit large changes in bathymetry across grid cells \citep{madec2012nemo}. {\color{red} Model constraints require that all passages be at least two grid points wide.  This restriction required choices in either widening or closing narrow straits.  In particular, Active Pass was closed and Porlier Pass was widened (Figure \ref{fig:zoom}).  Shallow regions were represented by constant depth of 4 m.  All of these compromises are expected to have significant effects locally but minimal effects on the large scale.}

@@

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results of a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid to give a two-dimensional forcing that varies with time.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. {\color{red} At the north, seasonal climatologies from \citet{thomson1981oceanography} were used.} %other references?

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. {\color{red}Tidal predictions were generated with a tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge. In order to prevent over-predicted constituents due to high surge years, the time series is first filtered by applying a Doodson filter \citep{parker2007NOAA} and removing periods with large non-tidal amplitudes. Long period constituents such as SA, SSA, etc are not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected. } The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

{\color{red} Although Tofino is not the closest tide guage to the western boundary, we have found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. }

\begin{table}[h]
\centering 
\tbl{{\color{red}Average surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the surge events discussed on this article. A negative delay indicates that the peak surge occurred later than at Tofino.}} 
{\begin{tabular}{|l |c | c|}
\hline 
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
\hline
Tofino  & 0.707 & -- \\
Toke Point & 1.17 & -2.83 \\
Neah Bay & 0.658 & -1.83 \\
Bamfield & 0.658 & -0.5 \\
Winter Harbour & 0.572 & 0.333 \\
Port Hardy & 0.491 & -2.67\\
\hline
\end{tabular}}
\label{tab:pacificsurge}
\end{table}

The tidal forcing and remote forcing were applied using the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}, a radiation scheme that allows disturbances to propagate out of the domain without reflections off of the open boundary. The barotropic velocity component of the remote forcing was set to zero. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4 (the version used here).  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. {\color{red} No boundary conditions were set on the baroclinic velocities at the north.}

@@

{\color{red} The ocean surface is forced with momentum fluxes, solar and non-solar heat fluxes, and precipitation from a 33-km global atmospheric reforecasting model suitable for use in ocean modelling \citep{smith2014new}.} Forecasts from the period of 2002-2012 are available. One additional simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the surface atmospheric pressure is also included in the ocean momentum equations \citep{madec2012nemo}. In the open ocean, the surface atmospheric pressure primarily induces an inverse barometer (IB) sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the acceleration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure (set to 101,000 Pa). Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. In coastal and shelf seas, the atmospheric pressure forcing can also induce significant non-IB sea surface height, as demonstrated in previous studies of the Hudson Bay and Labrador shelf system \citep{wright1987influence,young1995synoptic}.

River input provides a significant volume of fresh water to the Salish Sea and can influence stratification, circulation, and primary productivity. However, most rivers in the domain are not gauged so parametrizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages. Freshwater runoff from each watershed was divided among the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source from 0 to 3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface {\color{red}ocean temperature}. A total of 150 rivers were parametrized by this method and their locations are indicated by the green dots in Figure \ref{fig:domain}.
